Connecting R with D3 for dynamic graphics, to explore multivariate data with tours

### 2019-03-03 review_rebuttal.Rmd

---
title: "Review rebuttal"
author: "Michael Kipp, Ursula Laa, Di Cook"
date: "01/03/2019"
output: html_document
---

*We are pleased to accept your paper "Connecting R with D3 for dynamic graphics, to explore multivariate data with tours" subject to the major revisions described below.*

*Though the reviewer was critical, they thought there were enough interesting parts to form a paper around, though they expect that there will be some fleshing out, as described below:*

*> Overall recommendation: Accept with major revisions.*

*I'm very underwhelmed by this paper, to be honest. I think it's a bit of a tricky one to get excited about because all the Shiny experts will be reading thinking "well of course I knew you could do that" whereas all the people less familiar with Shiny won't really understand the approach. Other than that there's nothing fundamentally wrong with the approach, it's a sensible solution to a real problem and they've thoughtfully evaluated the pros and cons at the end.*

**Thanks, this summary is appreciated**

*> Overview: Is the article important to R the community? Is the approach sensible? Are technology and methods up-to-date?*

*Yes, it's sort of minimally important, yes. I think to make it feel more relevant it would be better to reframe as saying "hey, did you know you can run ANY JAVASCRIPT YOU LIKE in a Shiny application and you can pass WHATEVER YOU WANT INCLUDING JSON back and forth, this massively widens the scope of what you can do with Shiny, here's a really good example of what we did". It seems a bit churlish, I suppose, as well, but having said the tourrGui package is out of date it makes you think maybe they should turn theirs into a package*

**Additional sentences have been added to the introduction and abstract to indicate the general nature of the basic tool.**

*> Article: Is anything unclear? Is anything missing? Are the examples helpful? Is there sufficient background, including alternative approaches in R?**

*For the novices, it would be nice for them to include the method to run their code straight from the terminal:*

```
library(shiny)
runGitHub("makipp/TourrGuiD3")
```

**Thank you, a shiny app has been created and these instructions will run it.**

```
devtools::install_github("uschiLaa/tourrGUID3")
library(tourrGUID3)
launchApp(system.file("extdata", "geozoo.csv", package = "tourrGUID3"))
```

*I really think they need to make the paper more specialised to appeal to a Shiny audience, telling people who know Shiny stuff they don't already know, or they need to make it more general, filling in a few blanks for people who don't know Shiny. For example, they skip straight over the invalidateLater() function, they could at least tell us what it's doing.*

**More explanation has been added**

*They also say "The projections are calculated using the tour object in an observe() environment, which repeatedly runs the code until a reactive variable is changed, at which point it is “invalidated” and then re-started". I think this is a bit confusing. In fact the code ONLY reruns when it is invalidated (whether by invalidateLater() or a reactive value), that's how observe() works. I'm sure that's what they mean, really, but it isn't expressed very well.*

**Change made**

*I'm not over the moon about the dependency on ggvis, it's not that important since this is hardly production code, but this package is officially dormant and I think it's best avoided where possible, to save readers from going down a blind alley with it.*

**Dependency on ggvis removed, it was unnecessary, and yes, was simply left over from an older version used in testing.**

*The last thing is I would like them just to give one sentence to explain how the D3 actually ends up on the page. Really basic stuff for your Shiny savvy readers, but I don't think it is of much interest to them anyway because it's all basic stuff for the Shiny savvy.*

**check this**

*> Package: Could the functions presented be improved? Some ideas at http://bit.ly/1aYwkp3.*

*I think the code is fine. For me this paper could be improved by explicitly broadening the scope and hand holding a bit for people who are not totally au fait with Shiny.*

**check this**

*We would appreciate a revised version and point-by-point response to the reviewers comments within 2 months. Remember, that when responding to the reviewer's comments, your job is to persuade me, the editor, that either you've dealt with the issue, or that it's not relevant. To this end, please produce a single document that includes all the reviewers comments mingled with your responses. I particularly recommend the strategy described at http://matt.might.net/articles/peer-review-rebuttals/.*

### 2019-03-03 paper-tourrd3.Rmd
Some more details about D3 interface.

Dd you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth? This massively widens the scope of what you can do with Shiny, and generating a tour of multivariate data with this approach is a really good example of what is possible.

@@

@@

```
observeEvent(input$restart_random,
  {
    p <- length(input$variables)
    b <- matrix(runif(2*p), p, 2)
    rv$tour <- 
      new_tour(as.matrix(rv$d[input$variables]),
              choose_tour(input$type, 
              input$guidedIndex,
              c(rv$class[[1]]), 
              input$scagType), b)
})
```

The projections are calculated using the tour object in an `observe()` environment, which re-executes the code whenever it is invalidated. The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval. The projections are calculated using the following code block:

```
observe({
  tour <- rv$tour
  aps <- rv$aps
  step <- tour(aps / fps)
  if (!is.null(step)) {
    invalidateLater(1000 / fps)
    j <- center(rv$mat %*% step$proj)
    j <- cbind(j, class = rv$class)
    colnames(j) <- NULL
    session$sendCustomMessage(type = "data", 
      message = toJSON(j))
  }
  else {
    if (length(rv$mat[1, ]) < 3) {
    session$sendCustomMessage(type = "debug", 
      message = "Error: Need > 2 variables.")
    } else {
    session$sendCustomMessage(type = "debug", 
    message = "Guided tour finished: no better bases found.")
    }
  }
})
```

>>>>

Did you know you can run *any javascript you like* in a Shiny application and you can pass *whatever you want including JSON* back and forth? This massively widens the scope of what you can do with Shiny, and generating a tour of multivariate data with this approach is a really good example of what is possible.

@@

On the D3 side we can access the id defined in Shiny, and for example assign it to a scalable vector graphics (svg) object to be filled in D3 and rendered onto the Shiny app.

```
var svg = d3.select("#d3_output")
    .append("svg")
    .attr("width", w)
    .attr("height", h);
```

@@

```
observeEvent(input$restart_random,
  {
    p <- length(input$variables)
    b <- matrix(runif(2*p), p, 2)
    rv$tour <- 
      new_tour(as.matrix(rv$d[input$variables]),
              choose_tour(input$type, 
              input$guidedIndex,
              c(rv$class[[1]])), b)
})
```

The projections are calculated using the tour object in an `observe()` environment, which re-executes the code whenever it is invalidated. The invalidation is either by a change in reactive value inside the code block, or we can schedule a re-execution by explicitly invalidating the observer after a selected interval using `invalidateLater()`. The projections are calculated using the following code block:

```
observe({
      if (length(rv$mat[1, ]) < 3) {
        session$sendCustomMessage(type = "debug",
        message = "Error: Need >2 variables.")
      }
      aps <- rv$aps
      tour <- rv$tour
      step <- rv$tour(aps / fps)
      invalidateLater(1000 / fps)
      j <- center(rv$mat %*% step$proj)
      j <- cbind(j, class = rv$class)
      colnames(j) <- NULL
      session$sendCustomMessage(type = "data",
                                message =
                                list(d = toJSON(data.frame(pL=rv$pLabel[,1],x=j[,2],y=j[,1],c=j[,3])),
                                a = toJSON(data.frame(n=rv$vars,y=step$proj[,1],x=step$proj[,2]))))
})
```

### 2019-03-03 paper-tourrd3.tex
Some more details about D3 interface.

Dd you know you can run \emph{any javascript you like} in a Shiny
application and you can pass \emph{whatever you want including JSON}
back and forth? This massively widens the scope of what you can do with
Shiny, and generating a tour of multivariate data with this approach is
a really good example of what is possible.

@@

\begin{Schunk}
\begin{Sinput}
library(tourr)
# quartz() # to display on a Mac; X11() # For windows; The Rstudio graphics
# device is not advised
animate_dist(flea[, 1:6], center = TRUE)
\end{Sinput}
\end{Schunk}

@@

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. The list entries of the
``message'' can parsed in D3 by its \texttt{data()} method, e.g.
\texttt{data(message.d)} to access the projected data points, and we can
access each column through the column names assigned in the original
dataframe, and loop over all rows for rendering. All of the code
required to render the scatterplots and legends, along with colours, is
JavaScript code in the file \texttt{d3anim.js}. In particular, the data
from R is handled with the following code:

@@

\begin{verbatim}
observeEvent(input$restart_random,
  {
    p <- length(input$variables)
    b <- matrix(runif(2*p), p, 2)
    rv$tour <- 
      new_tour(as.matrix(rv$d[input$variables]),
              choose_tour(input$type, 
              input$guidedIndex,
              c(rv$class[[1]]), 
              input$scagType), b)
})
\end{verbatim}

The projections are calculated using the tour object in an
\texttt{observe()} environment, which re-executes the code whenever it
is invalidated. The invalidation is either by a change in reactive value
inside the code block, or we can schedule a re-execution by explicitly
invalidating the observer after a selected interval. The projections are
calculated using the following code block:

\begin{verbatim}
observe({
  tour <- rv$tour
  aps <- rv$aps
  step <- tour(aps / fps)
  if (!is.null(step)) {
    invalidateLater(1000 / fps)
    j <- center(rv$mat %*% step$proj)
    j <- cbind(j, class = rv$class)
    colnames(j) <- NULL
    session$sendCustomMessage(type = "data", 
      message = toJSON(j))
  }
  else {
    if (length(rv$mat[1, ]) < 3) {
    session$sendCustomMessage(type = "debug", 
      message = "Error: Need > 2 variables.")
    } else {
    session$sendCustomMessage(type = "debug", 
    message = "Guided tour finished: no better bases found.")
    }
  }
})
\end{verbatim}

>>>>

Did you know you can run \emph{any javascript you like} in a Shiny
application and you can pass \emph{whatever you want including JSON}
back and forth? This massively widens the scope of what you can do with
Shiny, and generating a tour of multivariate data with this approach is
a really good example of what is possible.

@@

\begin{Schunk}
\begin{Sinput}
library(tourr)
# quartz() # to display on a Mac;
# X11() # For windows;
# The Rstudio graphics device is not advised
animate_dist(flea[, 1:6], center = TRUE)
\end{Sinput}
\end{Schunk}

@@

On the D3 side we can access the id defined in Shiny, and for example
assign it to a scalable vector graphics (svg) object to be filled in D3
and rendered onto the Shiny app.

\begin{verbatim}
var svg = d3.select("#d3_output")
    .append("svg")
    .attr("width", w)
    .attr("height", h);
\end{verbatim}

@@

This code is from the observe environment from the \texttt{server.R}
file. It converts the matrix of projected data points to JSON format,
and sends it to JavaScript with the id data. The list entries of the
``message'' can parsed in D3 by its \texttt{data()} method,
e.g.~\texttt{data(message.d)} to access the projected data points, and
we can access each column through the column names assigned in the
original dataframe, and loop over all rows for rendering. All of the
code required to render the scatterplots and legends, along with
colours, is JavaScript code in the file \texttt{d3anim.js}. In
particular, the data from R is handled with the following code:

@@

\begin{verbatim}
observeEvent(input$restart_random,
  {
    p <- length(input$variables)
    b <- matrix(runif(2*p), p, 2)
    rv$tour <- 
      new_tour(as.matrix(rv$d[input$variables]),
              choose_tour(input$type, 
              input$guidedIndex,
              c(rv$class[[1]])), b)
})
\end{verbatim}

The projections are calculated using the tour object in an
\texttt{observe()} environment, which re-executes the code whenever it
is invalidated. The invalidation is either by a change in reactive value
inside the code block, or we can schedule a re-execution by explicitly
invalidating the observer after a selected interval using
\texttt{invalidateLater()}. The projections are calculated using the
following code block:

\begin{verbatim}
observe({
      if (length(rv$mat[1, ]) < 3) {
        session$sendCustomMessage(type = "debug",
        message = "Error: Need >2 variables.")
      }
      aps <- rv$aps
      tour <- rv$tour
      step <- rv$tour(aps / fps)
      invalidateLater(1000 / fps)
      j <- center(rv$mat %*% step$proj)
      j <- cbind(j, class = rv$class)
      colnames(j) <- NULL
      session$sendCustomMessage(type = "data",
                                message =
                                list(d = toJSON(data.frame(pL=rv$pLabel[,1],x=j[,2],y=j[,1],c=j[,3])),
                                a = toJSON(data.frame(n=rv$vars,y=step$proj[,1],x=step$proj[,2]))))
})
\end{verbatim}
